---
title: "Group_21_Analysis"
author: "group21"
format:
  pdf:
    geometry: margin=1in
    fontsize: 11pt
editor: visual
---

## Introduction

Spam emails are a common problem in email communication and can affect both productivity and security. In this project, we use a spam email dataset to explore which text features are associated with an email being classified as spam. To do this, we fit a binomial logistic regression model and evaluate how well these variables help distinguish spam from non-spam emails.

## Data preprocessing

```{r}
#| label: setup
#| include: false
#| warning: false
#| message: false

# Hide warnings and package messages in the final output
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)

# Load packages used in the analysis
library(tidyverse)
library(broom)
library(caret)
library(pROC)
library(car)
library(knitr)

# Set seed so the train/test split can be reproduced
set.seed(21)

# Colours used in plots
accent_col <- "#0B3C5D"
highlight_col <- "#E76F51"

# Try to find the folder where the qmd file is located
# If that does not work, use the current working directory
project_dir <- tryCatch({
  in_dir <- knitr::current_input(dir = TRUE)
  if (nzchar(in_dir)) in_dir else getwd()
}, error = function(e) {
  getwd()
})

# Build the path to the data file
data_path <- file.path(project_dir, "dataset21.csv")

# Fallback option if the file is already in the working directory
if (!file.exists(data_path)) {
  data_path <- "dataset21.csv"
}

# Read the dataset
df <- read.csv(data_path)

# Simple helper to turn model output into a clean table
coef_table_fn <- function(model) {
  tidy(model, conf.int = TRUE, exponentiate = TRUE) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      term = str_replace_all(term, "_", " "),
      estimate = round(estimate, 3),
      conf.low = round(conf.low, 3),
      conf.high = round(conf.high, 3),
      p.value = signif(p.value, 3)
    ) %>%
    transmute(
      Predictor = term,
      `Odds ratio` = estimate,
      `95% CI` = paste0(conf.low, " to ", conf.high),
      `p-value` = p.value
    )
}

# Simple helper to collect the main model performance measures
perf_table_fn <- function(actual, pred_prob) {
  pred_class <- factor(ifelse(pred_prob > 0.5, "y", "n"), levels = c("n", "y"))
  cm <- confusionMatrix(pred_class, actual, positive = "y")
  roc_obj <- roc(actual, pred_prob, levels = c("n", "y"), quiet = TRUE)

  tibble(
    Accuracy = as.numeric(cm$overall["Accuracy"]),
    Sensitivity = as.numeric(cm$byClass["Sensitivity"]),
    Specificity = as.numeric(cm$byClass["Specificity"]),
    AUC = as.numeric(auc(roc_obj))
  ) %>%
    mutate(across(everything(), \(x) round(x, 4)))
}
```

```{r}
#| label: check-data

# Look at the structure of the data
glimpse(df)

# Check if there are any missing values
any(is.na(df))
```

There are no missing values in the entire dataset.

## Exploratory analysis

```{r}
ggplot(df, aes(x = crl.tot)) +
  geom_histogram(
    bins = 30,
    fill = "#0B3C5D",
    color = "white"
  ) +
  labs(
    title = "Histogram of crl.tot",
    x = "crl.tot",
    y = "Frequency"
  ) +
  theme_minimal()

```

The variable `crl.tot` was strongly right-skewed, so we used `log(1 + crl.tot)` to reduce skewness and improve the fit of the GLM.

```{r}

#| label: prepare-data
#| include: false

# Make sure the response variable is treated as a factor
# n = non-spam, y = spam
df <- df %>%
  mutate(
    yesno = factor(yesno, levels = c("n", "y"))
  )

# Create log-transformed versions of skewed predictors
# log1p(x) means log(1 + x)
df <- df %>%
  mutate(
    log_crl_tot = log1p(crl.tot),
    log_dollar  = log1p(dollar),
    log_bang    = log1p(bang),
    log_money   = log1p(money),
    log_n000    = log1p(n000),
    log_make    = log1p(make)
  )

```

Before modelling, we prepared the dataset by converting the response variable `yesno` to a factor and creating a log-transformed version of `crl.tot` to reduce right skewness. We also inspected the class balance between spam and non-spam emails.

The dataset contains 557 non-spam emails (60.5%) and 363 spam emails (39.5%).

```{r}
ggplot(df, aes(x = log_crl_tot)) +
  geom_histogram(
    bins = 30,
    fill = "#0B3C5D",
    color = "white"
  ) +
  labs(
    title = "Histogram of log(1 + crl.tot)",
    x = "log(1 + crl.tot)",
    y = "Frequency"
  ) +
  theme_minimal()
```

We drew a histogram for the transformed variable and found that it is much less skewed, making it more suitable for modeling.

```{r}

num_vars <- c("crl.tot", "dollar", "bang", "money", "n000", "make")

for (v in num_vars) {
  print(
    ggplot(df, aes(x = .data[[v]])) +
      geom_histogram(
        bins = 30,
        fill = "#0B3C5D",
        color = "white"
      ) +
      labs(
        title = paste("Histogram of", v),
        x = v,
        y = "Frequency"
      ) +
      theme_minimal()
  )
}
```

Most predictors are right-skewed, which is expected for word- and symbol-frequency variables. Among them, `crl.tot` shows particularly strong skewness, but the same pattern is also visible for several other predictors.

```{r}

for (v in num_vars) {
  print(
    ggplot(df, aes(x = yesno, y = .data[[v]], fill = yesno)) +
      geom_boxplot(show.legend = FALSE) +
      labs(
        title = paste("Boxplot of", v, "by Email Class"),
        x = "Email Class",
        y = v
      ) +
      theme_minimal()
  )
}
ggplot(df, aes(x = yesno, y = log_crl_tot, fill = yesno)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Boxplot of log(1 + crl.tot) by Email Class",
    x = "Email Class",
    y = "log(1 + crl.tot)"
  ) +
  theme_minimal()
```

These box plots illustrate that the above-mentioned variables occur more frequently in spam emails than in normal emails. We did not remove the outliers, because they are likely to be genuine features of spam emails rather than errors in the data.

```{r}
# Summary table for the dataset
data_summary <- tibble(
  Metric = c("Rows", "Columns", "Missing values", "Non-spam emails", "Spam emails"),
  Value = c(nrow(df), ncol(df), sum(is.na(df)), sum(df$yesno == "n"), sum(df$yesno == "y"))
)

variable_summary <- df %>%
  select(crl.tot, dollar, bang, money, n000, make) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(
    Mean = round(mean(Value), 3),
    Median = round(median(Value), 3),
    `Std. Dev.` = round(sd(Value), 3),
    Max = round(max(Value), 3),
    .groups = "drop"
  )

knitr::kable(data_summary, caption = "Basic dataset summary")
knitr::kable(variable_summary, caption = "Summary of numeric predictors")
```

## Modelling

We fitted a generalized linear model (GLM) with a binomial family and logit link, that is, a logistic regression model. This model is appropriate because the response variable `yesno` is binary, indicating whether an email is spam or not spam. The model estimates how the predictor variables affect the probability that an email is classified as spam.

$$
\log \left( \frac{P(\text{spam})}{1 - P(\text{spam})} \right)
= \beta_0 + \beta_1 \,\text{log\_crl\_tot} + \beta_2 \,\text{dollar} + \beta_3 \,\text{bang} + \beta_4 \,\text{money} + \beta_5 \,\text{n000} + \beta_6 \,\text{make}.
$$

```{r}
# Fit the full logistic regression model
mod1 <- glm(
  yesno ~ log_crl_tot + dollar + bang + money + n000 + make,
  data = df,
  family = binomial
)

mod1_coef_table <- coef_table_fn(mod1)
knitr::kable(mod1_coef_table, caption = "Coefficient summary for the full model")
```

The GLM results show that `log_crl_tot`, `dollar`, `bang`, `money`, and `n000` are all significant positive predictors of spam, meaning that higher values of these variables are associated with a higher probability that an email is classified as spam. In contrast, `make` is not statistically significant, suggesting that it does not provide additional predictive value once the other variables are included in the model. Overall, the model indicates that features such as capital-letter runs, dollar signs, exclamation marks, the word “money”, and the string “000” are important indicators of spam.

```{r}
# Get fitted probabilities and summarise model performance
df$pred_prob <- predict(mod1, type = "response")
mod1_perf_table <- perf_table_fn(df$yesno, df$pred_prob)
knitr::kable(mod1_perf_table, caption = "In-sample performance of the full model")
```

The GLM achieved an in-sample accuracy of 85.11%, which is substantially higher than the no-information rate of 60.54%. The model has high specificity (94.61%), meaning it is very good at identifying non-spam emails, and moderate-to-strong sensitivity (70.52%), meaning it also detects most spam emails. Overall, the model performs well, although some spam emails are still missed. However, because this accuracy is calculated on the same data used to fit the model, it is likely to be somewhat optimistic.

```{r}
# Plot the ROC curve for the full model
roc_obj <- roc(response = df$yesno, predictor = df$pred_prob, levels = c("n", "y"))
ggroc(roc_obj, linewidth = 1.2, color = accent_col) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  annotate(
    "label",
    x = 0.72,
    y = 0.16,
    label = paste0("AUC = ", round(as.numeric(auc(roc_obj)), 3)),
    fill = "white",
    color = accent_col,
    linewidth = 0.25
  ) +
  labs(
    title = "ROC Curve for the Full Model",
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal()
```

The ROC curve lies well above the diagonal reference line, indicating that the GLM has good discriminatory ability in distinguishing spam from non-spam emails. This suggests that the model performs substantially better than random classification and provides strong overall predictive performance.

## Reduced Model

Since `make` was not statistically significant in the full model, we fitted a reduced model excluding `make` and compared the two models. If the reduced model shows similar or better fit, it would be preferred for parsimony.

```{r}
# Fit the reduced model without `make`
mod2 <- glm(
  yesno ~ log_crl_tot + dollar + bang + money + n000,
  data = df,
  family = binomial
)

df$pred_prob2 <- predict(mod2, type = "response")

model_compare_table <- bind_rows(
  perf_table_fn(df$yesno, df$pred_prob) %>% mutate(Model = "Full model", AIC = round(AIC(mod1), 4)),
  perf_table_fn(df$yesno, df$pred_prob2) %>% mutate(Model = "Reduced model", AIC = round(AIC(mod2), 4))
) %>%
  select(Model, AIC, Accuracy, Sensitivity, Specificity, AUC)

lrt_result <- anova(mod1, mod2, test = "Chisq")
lrt_table <- tibble(
  Comparison = "Full vs Reduced",
  `Deviance difference` = round(abs(lrt_result$Deviance[2]), 4),
  `p-value` = signif(lrt_result$`Pr(>Chi)`[2], 3)
)

knitr::kable(coef_table_fn(mod2), caption = "Coefficient summary for the reduced model")
knitr::kable(model_compare_table, caption = "Comparison of the full and reduced models")
knitr::kable(lrt_table, caption = "Nested model test")

model_compare_plot <- model_compare_table %>%
  select(Model, Accuracy, Sensitivity, Specificity, AUC) %>%
  pivot_longer(
    -Model,
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(model_compare_plot, aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.65) +
  scale_fill_manual(
    values = c(
      Accuracy = accent_col,
      Sensitivity = "#2A9D8F",
      Specificity = "#E9C46A",
      AUC = highlight_col
    )
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Performance Comparison: Full vs Reduced Model",
    x = NULL,
    y = "Score",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

The reduced model has a slightly lower AIC than the full model, suggesting a marginally better fit with fewer predictors. Its classification performance remains almost unchanged, with an accuracy of 85.0%, indicating that removing `make` does not meaningfully reduce predictive performance. Therefore, the reduced model may be preferred for parsimony.

## Model limitations and improvement

Although the original and reduced models are useful, they still have two main limitations. First, only `crl.tot` was transformed, even though the other frequency variables are also strongly right-skewed. Second, model performance was assessed on the same data used for fitting, which tends to overestimate predictive accuracy. To improve the model without changing the overall structure of the report too much, we fit one additional logistic regression using transformed predictors and evaluate it on a held-out test set.

The final improved model is:

$$
\log \left( \frac{P(\text{spam})}{1 - P(\text{spam})} \right)
= \beta_0 + \beta_1 \,\text{log\_crl\_tot} + \beta_2 \,\text{log\_dollar}
+ \beta_3 \,\text{log\_bang} + \beta_4 \,\text{log\_money}
+ \beta_5 \,\text{log\_n000}.
$$

```{r}
# Split the data into training and test sets
train_index <- createDataPartition(df$yesno, p = 0.80, list = FALSE)
train_df <- df[train_index, ]
test_df <- df[-train_index, ]

# Fit the improved model using transformed predictors
mod3 <- glm(
  yesno ~ log_crl_tot + log_dollar + log_bang + log_money + log_n000,
  data = train_df,
  family = binomial
)

knitr::kable(coef_table_fn(mod3), caption = "Coefficient summary for the improved model")

coef_plot_df <- tidy(mod3, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = factor(
      term,
      levels = c("log_crl_tot", "log_dollar", "log_bang", "log_money", "log_n000"),
      labels = c("log crl tot", "log dollar", "log bang", "log money", "log n000")
    )
  )

ggplot(coef_plot_df, aes(x = estimate, y = forcats::fct_rev(term))) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray60") +
  geom_errorbar(
    aes(xmin = conf.low, xmax = conf.high),
    width = 0.18,
    color = accent_col
  ) +
  geom_point(size = 3, color = highlight_col) +
  scale_x_log10() +
  labs(
    title = "Odds Ratios for the Improved Model",
    x = "Odds ratio (log scale)",
    y = NULL
  ) +
  theme_minimal()
```

This improved model keeps the same logistic regression framework, but uses transformed versions of the main skewed predictors throughout and excludes `make`, which was not useful in the earlier models.

```{r}
# Evaluate the improved model on the test set
test_prob <- predict(mod3, newdata = test_df, type = "response")
test_class <- factor(ifelse(test_prob > 0.5, "y", "n"), levels = c("n", "y"))
test_cm <- confusionMatrix(test_class, test_df$yesno, positive = "y")
test_auc <- auc(roc(test_df$yesno, test_prob, levels = c("n", "y"), quiet = TRUE))

performance_table <- tibble(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "AUC"),
  Value = round(
    c(
      as.numeric(test_cm$overall["Accuracy"]),
      as.numeric(test_cm$byClass["Sensitivity"]),
      as.numeric(test_cm$byClass["Specificity"]),
      as.numeric(test_auc)
    ),
    4
  )
)

knitr::kable(
  performance_table,
  caption = "Performance of the improved model on the test set"
)

improvement_compare_table <- bind_rows(
  perf_table_fn(df$yesno, df$pred_prob) %>%
    mutate(Model = "Original full model", Evaluation = "In-sample", AIC = round(AIC(mod1), 4)),
  perf_table_fn(df$yesno, df$pred_prob2) %>%
    mutate(Model = "Original reduced model", Evaluation = "In-sample", AIC = round(AIC(mod2), 4)),
  perf_table_fn(test_df$yesno, test_prob) %>%
    mutate(Model = "Improved test-set model", Evaluation = "Test set", AIC = round(AIC(mod3), 4))
) %>%
  select(Model, Evaluation, AIC, Accuracy, Sensitivity, Specificity, AUC)

knitr::kable(
  improvement_compare_table,
  caption = "Combined performance summary across models"
)

improvement_compare_plot <- improvement_compare_table %>%
  pivot_longer(
    c(Accuracy, Sensitivity, Specificity, AUC),
    names_to = "Metric",
    values_to = "Value"
  )

ggplot(improvement_compare_plot, aes(x = Model, y = Value, fill = Metric)) +
  geom_col(position = position_dodge(width = 0.72), width = 0.65) +
  scale_fill_manual(
    values = c(
      Accuracy = accent_col,
      Sensitivity = "#2A9D8F",
      Specificity = "#E9C46A",
      AUC = highlight_col
    )
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(
    title = "Combined Performance Across the Three Models",
    x = NULL,
    y = "Score",
    fill = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

On the held-out test set, the improved model achieves an accuracy of about 82.0%, with sensitivity of about 70.8%, specificity of about 89.2%, and an AUC of about 0.888. These values are slightly lower than the earlier in-sample results, but they provide a more realistic measure of how the model would perform on new emails.

```{r}
# Plot the ROC curve for the improved model
roc_obj2 <- roc(response = test_df$yesno, predictor = test_prob, levels = c("n", "y"))
ggroc(roc_obj2, linewidth = 1.2, color = highlight_col) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray60") +
  annotate(
    "label",
    x = 0.72,
    y = 0.16,
    label = paste0("AUC = ", round(as.numeric(auc(roc_obj2)), 3)),
    fill = "white",
    color = highlight_col,
    linewidth = 0.25
  ) +
  labs(
    title = "ROC Curve for the Improved Model on the Test Set",
    x = "1 - Specificity",
    y = "Sensitivity"
  ) +
  theme_minimal()
```

The test-set ROC curve remains well above the diagonal line, showing that the improved model still has strong discriminatory ability even under more realistic evaluation.

## Conclusion

In this project, we first prepared the data by converting `yesno` into a factor and creating a log-transformed version of `crl.tot`. From the exploratory data analysis, we found that most variables were right-skewed, especially `crl.tot`, so transforming skewed predictors was appropriate. We then fitted a binomial logistic regression model and found that `log_crl_tot`, `dollar`, `bang`, `money`, and `n000` were important positive predictors of spam, while `make` was not significant. After removing `make`, the reduced model still performed well.

To make the analysis more reliable, we also fitted an improved logistic regression using transformed versions of the main skewed predictors and evaluated it on a held-out test set. This gave a more realistic accuracy of about 82%, while still showing strong classification performance. Overall, the results suggest that emails containing more capital-letter runs, dollar signs, exclamation marks, the word “money”, and the string “000” are more likely to be classified as spam.
